A Dependently Typed Multi-Stage Calculus\thanks{Supported by organization x.}

### 2019-05-22 main.tex
Wrote type equivalence.

    We develop yet another typed multi-stage calculus \LMD.
    It extends Hanada and Igarashi's \LTP with dependent types.

    A multi-stage calculus enables us to generate and execute codes at runtime.
    It can improve the performance of programs by generating optimized codes for given inputs.

    Dependent types are types dependent on values. 
    A vector with its length is a famous example of dependent types and it enables us to omit boundary checking.

    In this paper, we design \LMD by introducing dependent type into $\lambda^{\TW\%}$.
    You can make more efficient programs from existing dependent typed programs with \LMD.

    \LMD has a simple, substitution-based full-reduction semantics and enjoys basic properties of subject reduction, confluence, and strong normalization, and progress.
    It also includes an evaluation context which satisfies unique decomposition.

    The main technical points of this paper are how to deal with Cross Stage Persistence of multi-stage calculuses which allows using a value in quoted code in a dependent type system.
    Especially, the way of handling CSP in equivalence rules of dependent types wasn't clear.
    In this paper, we give reasonable equivalence rules to handle them.

@@

% \QCsp以外の説明

@@

\begin{figure}
	\begin{center}
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau@A \V \rho \E \pi :: *@A}{\G\V\Pi x:\tau.\rho \E \Pi x:\sigma.\pi :: *@A}{\QTAbs} \andalso
		\infrule{\G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso \G\V M \E N : \rho @A}{\G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A}{\QTApp} \\[2mm]
		\infrule{\G\V \tau \E \sigma :: *@{A\alpha}}{\G\V \TW_{\alpha} \tau \E \TW_{\alpha} \sigma :: *@A}{\textsc{QT-$\TW$}}\andalso
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V \forall\alpha.\tau \E  \forall\alpha.\sigma :: *@A}{\textsc{QT-Gen}} \andalso
		\infrule{\G\V \tau \E \sigma :: K@A}{\G\V \tau \E \sigma :: K@{A\alpha}}{\textsc{QT-Csp}} \\[4mm]
		\infrule{\G\V \tau::K@A}{\G\V \tau\E\tau :: K@A}{\textsc{QT-Refl}} \andalso
		\infrule{\G\V \tau \E \sigma :: K@A}{\G\V \sigma \E \tau :: K@A}{\textsc{QT-Sym}} \andalso
		\infrule{\G\V \tau \E \sigma :: K@A \andalso \G\V \sigma \E \rho  :: K@A}{\G\V \tau \E \rho  :: K@A}{\textsc{QT-Trans}} \\[4mm]
		\caption{Type Equivalence Rules}
		\label{fig:type-equivalence-rules}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau@A \V M \E N : \rho @A}{\G\V\lambda x:\tau.M \E \lambda x:\sigma.N : (\Pi x:\tau.\rho)@A}{\QAbs} \andalso
		\infrule{\G\V M \E L : (\Pi x:\sigma.\tau)@A \andalso \G\V N \E O : \sigma@A}{\G\V M\ N \E L\ O : \tau[x \mapsto N]@A}{\QApp} \\[2mm]
		\infrule{\G\V M \E N : \tau@{A\alpha}}{\G\V \TB_\alpha M \E \TB_\alpha N : \TW_\alpha \tau@A}{\QTB} \andalso
		\infrule{\G\V M \E N : \TW_\alpha \tau@A}{\G\V \TBL_\alpha M \E \TBL_\alpha N : \tau@{A\alpha}}{\QTBL} \\[2mm]
		\infrule{\G\V M\E N : \tau@A \andalso \alpha \notin \FTV(\G)\cup\FTV(A)}{\G\V \Lambda\alpha.M \E \Lambda\alpha.N : \forall\alpha.\tau@A}{\QGen} \andalso
		\infrule{\G\V M \E N:\forall\alpha.\tau@A}{\G\V M\ \epsilon \E N\ \epsilon : \tau[\alpha \mapsto \epsilon]@A}{\QIns}\andalso
		\infrule{\G\V M \E N : \tau @A}{\G\V\%_\alpha M \E \%_\alpha N : \tau@{A\alpha}}{\QCsp} \\[4mm]
		\infrule{\G\V M:\tau@A}{\G\V M\E M : \tau@A}{\QRefl} \andalso
		\infrule{\G\V M\E N : \tau@A}{\G\V N\E M : \tau@A}{\QSym} \andalso
		\infrule{\G\V M\E N : \tau@A \andalso \G\V N\E L : \tau@A}{\G\V M\E L : \tau@A}{\QTrans} \\[4mm]
		\infrule{\G,x:\sigma@A\V M:\tau@A \andalso \G\V N:\sigma@A}{\G\V(\lambda x:\sigma.M)\ N\E M[x\mapsto N] : \tau[x \mapsto N]@A}{\QBeta} \andalso
		\infrule{\G\V M:(\Pi x:\sigma.\tau)@A \andalso x\notin \text{FV}(M)}{\G\V(\lambda x:\sigma.M\ x)\E M: (\Pi x:\sigma.\tau)@A}{\QEta} \\[2mm]
		\infrule{\G\V M \E N : \tau@A}{\G\V \TBL_\alpha(\TB_\alpha M) \E N : \tau @A}{\QTBLTB} \andalso
		\infrule{\G\V (\Lambda\alpha.M) : \forall\alpha.\tau@A}{\G\V (\Lambda\alpha.M)\ \epsilon \E M[\alpha \mapsto \epsilon] : \tau[\alpha \mapsto \epsilon]@A}{\QLambda} \\[2mm]
		\infrule{\G\V M:\tau@{A\alpha} \andalso \G\V M:\tau@A}{\G\V\%_\alpha M \E M : \tau@{A\alpha}}{\QPercent} \andalso
		\caption{Tem Equivalence Rules}
		\label{fig:tem-equivalence-rules}
	\end{center}
\end{figure}

>>>>

    We develop yet another typed multi-stage calculus \LMD.
    It extends Hanada and Igarashi's \LTP with dependent types.
    A multi-stage calculus enables us to generate and execute codes at runtime.
    It can improve the performance of programs by generating optimized codes for given inputs.
    Dependent types are types dependent on values. 
    A vector with its length is a famous example of dependent types and it enables us to omit boundary checking.
    In this paper, we design \LMD by introducing dependent type into $\lambda^{\TW\%}$.
    You can make more efficient programs from existing dependent typed programs with \LMD.
    \LMD has a simple, substitution-based full-reduction semantics and enjoys basic properties of subject reduction, confluence, and strong normalization, and progress.
    It also includes an evaluation context which satisfies unique decomposition.
    The main technical points of this paper are how to deal with Cross Stage Persistence of multi-stage calculuses which allows using a value in quoted code in a dependent type system.
    Especially, the way of handling CSP in equivalence rules of dependent types wasn't clear.
    In this paper, we give reasonable equivalence rules to handle them.

@@

% \QCsp以外の説明
We show type equivalence rules in Figure \ref{fig:type-equivalence-rules}.
All rules except \QTRefl, \QTSym, \QTTrans, and \QTApp\ are generated naturally from the typing rules.
\QTRefl, \QTSym, \QTTrans\ exist in order to make the type equivalence relation an equivalence relationship.
The rule \QTApp means that if there are two equivalent $\Pi$ type and two equivalent terms,
the result of applications are also equivalent.
A judgement of $\G \V M \E N : \rho @ A$ means that $M$ and $N$ are equivalent.

\begin{figure}
	\begin{center}
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau@A \V \rho \E \pi :: *@A}{\G\V\Pi x:\tau.\rho \E \Pi x:\sigma.\pi :: *@A}{\QTAbs} \\[2mm]
		\infrule{\G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso \G\V M \E N : \rho @A}{\G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A}{\QTApp} \\[2mm]
		\infrule{\G\V \tau \E \sigma :: *@{A\alpha}}{\G\V \TW_{\alpha} \tau \E \TW_{\alpha} \sigma :: *@A}{\textsc{QT-$\TW$}}\andalso
		\infrule{\G\V \tau \E \sigma :: K@A}{\G\V \tau \E \sigma :: K@{A\alpha}}{\textsc{QT-Csp}} \\[2mm]
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V \forall\alpha.\tau \E  \forall\alpha.\sigma :: *@A}{\textsc{QT-Gen}} \\[2mm]
		\infrule{\G\V \tau::K@A}{\G\V \tau\E\tau :: K@A}{\textsc{QT-Refl}} \andalso
		\infrule{\G\V \tau \E \sigma :: K@A}{\G\V \sigma \E \tau :: K@A}{\textsc{QT-Sym}} \\[2mm]
		\infrule{\G\V \tau \E \sigma :: K@A \andalso \G\V \sigma \E \rho  :: K@A}{\G\V \tau \E \rho  :: K@A}{\textsc{QT-Trans}}
		\caption{Type Equivalence Rules}
		\label{fig:type-equivalence-rules}
	\end{center}
\end{figure}

The equivalence of terms is defined in Figure \ref{fig:term-equivalence-rules}.
\QRefl, \QSym, \QTrans\ exist in order to make the term equivalence relation an equivalence relationship as with the type equivalence.

@@

\begin{figure}
	\begin{center}
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau@A \V M \E N : \rho @A}{\G\V\lambda x:\tau.M \E \lambda x:\sigma.N : (\Pi x:\tau.\rho)@A}{\QAbs} \andalso
		\infrule{\G\V M \E L : (\Pi x:\sigma.\tau)@A \andalso \G\V N \E O : \sigma@A}{\G\V M\ N \E L\ O : \tau[x \mapsto N]@A}{\QApp} \\[2mm]
		\infrule{\G\V M \E N : \tau@{A\alpha}}{\G\V \TB_\alpha M \E \TB_\alpha N : \TW_\alpha \tau@A}{\QTB} \andalso
		\infrule{\G\V M \E N : \TW_\alpha \tau@A}{\G\V \TBL_\alpha M \E \TBL_\alpha N : \tau@{A\alpha}}{\QTBL} \\[2mm]
		\infrule{\G\V M\E N : \tau@A \andalso \alpha \notin \FTV(\G)\cup\FTV(A)}{\G\V \Lambda\alpha.M \E \Lambda\alpha.N : \forall\alpha.\tau@A}{\QGen} \andalso
		\infrule{\G\V M \E N:\forall\alpha.\tau@A}{\G\V M\ \epsilon \E N\ \epsilon : \tau[\alpha \mapsto \epsilon]@A}{\QIns}\andalso
		\infrule{\G\V M \E N : \tau @A}{\G\V\%_\alpha M \E \%_\alpha N : \tau@{A\alpha}}{\QCsp} \\[4mm]
		\infrule{\G\V M:\tau@A}{\G\V M\E M : \tau@A}{\QRefl} \andalso
		\infrule{\G\V M\E N : \tau@A}{\G\V N\E M : \tau@A}{\QSym} \andalso
		\infrule{\G\V M\E N : \tau@A \andalso \G\V N\E L : \tau@A}{\G\V M\E L : \tau@A}{\QTrans} \\[4mm]
		\infrule{\G,x:\sigma@A\V M:\tau@A \andalso \G\V N:\sigma@A}{\G\V(\lambda x:\sigma.M)\ N\E M[x\mapsto N] : \tau[x \mapsto N]@A}{\QBeta} \andalso
		\infrule{\G\V M:(\Pi x:\sigma.\tau)@A \andalso x\notin \text{FV}(M)}{\G\V(\lambda x:\sigma.M\ x)\E M: (\Pi x:\sigma.\tau)@A}{\QEta} \\[2mm]
		\infrule{\G\V M \E N : \tau@A}{\G\V \TBL_\alpha(\TB_\alpha M) \E N : \tau @A}{\QTBLTB} \andalso
		\infrule{\G\V (\Lambda\alpha.M) : \forall\alpha.\tau@A}{\G\V (\Lambda\alpha.M)\ \epsilon \E M[\alpha \mapsto \epsilon] : \tau[\alpha \mapsto \epsilon]@A}{\QLambda} \\[2mm]
		\infrule{\G\V M:\tau@{A\alpha} \andalso \G\V M:\tau@A}{\G\V\%_\alpha M \E M : \tau@{A\alpha}}{\QPercent} \andalso
		\caption{Tem Equivalence Rules}
		\label{fig:term-equivalence-rules}
	\end{center}
\end{figure}
